\Title

### 2013-03-12 supplement_body.tex
remove packages disliked by plos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7.pdf}
\caption{Structure of viral populations and patient selection.
Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
seroconversion, from blue to red). Panel B) shows allele frequency trajectories for nonsynonymous
changes in the same patient. Here, the blue to red color map corresponds to the
position of the allele in \env{} from 5' to 3'. Panels C) and D) show analogous
plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008}) were excluded
from our analysis.}
\label{fig:aftp}
\end{center}
\end{figure}

\newpage
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome.pdf}
\caption{
Synonymous diversity across the HIV genome, as quantified by the normalized
codon entropy among sequences coding for the consensus amino acid. In most
parts of the genome, synonymous sites show little conservation. The synonymous
diversity peaks at the variable regions in {\it env} and is reduced in regions 
under purifying selection (RRE hairpin, second {\it tat}/{\it rev} exons). The
normalized codon entropy is calculated as follows (see the script
\texttt{codon\_entropy\_synonymous\_subtypeB.py} for the full algorithm): (i)
from a subtype B multiple sequence alignment (MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
consensus amino acid at each position in the HIV genome; (ii) we count how often
each codon coding for the consensus amino acid appears in the MSA; (iii) at each
amino acid position, we divide by the number of sequences in the MSA that had
the consensus amino acid at that position, obtaining {\it codon frequencies}
$\nu_c$; (iv) we calculate the codon entropy from each position as: $S := -
\sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for the
consensus amino acid at this site; (v) we divide by the maximal codon entropy of
that amino acid (e.g. $\log 2$ for twofold degenerate codons). All parts of
{\it env} that are part of a different gene (signaling peptide, second {\it rev}
exon) have been excluded from our main analysis, to avoid contamination by
protein selection in a different reading frame.
Note that all gap-rich columns of the MSA are stripped from this figure, so genes such as {\it env} might appear shorter than they actually
are.
}
\label{fig:syndiv_genome}
\end{center}
\end{figure}
\newpage
% 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Nonsynonymous changes outside of variable regions are deleterious}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[h]
% \begin{center}
% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
% {\it gag} in the LANL reference panel (filtered
% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
% of a certain type over the number of possible mutants is plotted. Nonsynonymous
% changes are observed less often, {\it ergo} are more conserved, than
% synonymous changes. It can be therefore assumed, as mentioned in the main text,
% that non-escape nonsynonymous changes involve a large fitness cost.}
% \label{fig:synnonsyncons}
% \end{center}
% \end{figure}
% \newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain.pdf}
\caption{
Time-dependent selection reduces fixation of nonsynonymous mutations. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\cite{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous mutations is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7.pdf}
\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7.pdf}
\caption{Structure of viral populations and patient selection.
Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
seroconversion, from blue to red). Panel B) shows allele frequency trajectories for nonsynonymous
changes in the same patient. Here, the blue to red color map corresponds to the
position of the allele in \env{} from 5' to 3'. Panels C) and D) show analogous
plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\cite{shankarappa_consistent_1999} and
ACH19542 and ACH19768 from ref.~\cite{bunnik_autologous_2008}) were excluded
from our analysis.}
\label{fig:aftp}
\end{center}
\end{figure}

\newpage
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome.pdf}
\caption{
Synonymous diversity across the HIV genome, as quantified by the normalized
codon entropy among sequences coding for the consensus amino acid. In most
parts of the genome, synonymous sites show little conservation. The synonymous
diversity peaks at the variable regions in {\it env} and is reduced in regions 
under purifying selection (RRE hairpin, second {\it tat}/{\it rev} exons). The
normalized codon entropy is calculated as follows (see the script
\texttt{codon\_entropy\_synonymous\_subtypeB.py} for the full algorithm): (i)
from a subtype B multiple sequence alignment (MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
consensus amino acid at each position in the HIV genome; (ii) we count how often
each codon coding for the consensus amino acid appears in the MSA; (iii) at each
amino acid position, we divide by the number of sequences in the MSA that had
the consensus amino acid at that position, obtaining {\it codon frequencies}
$\nu_c$; (iv) we calculate the codon entropy from each position as: $S := -
\sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for the
consensus amino acid at this site; (v) we divide by the maximal codon entropy of
that amino acid (e.g. $\log 2$ for twofold degenerate codons). All parts of
{\it env} that are part of a different gene (signaling peptide, second {\it rev}
exon) have been excluded from our main analysis, to avoid contamination by
protein selection in a different reading frame.
Note that all gap-rich columns of the MSA are stripped from this figure, so genes such as {\it env} might appear shorter than they actually
are.
}
\label{fig:syndiv_genome}
\end{center}
\end{figure}
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain.pdf}
\caption{
Time-dependent selection reduces fixation of nonsynonymous mutations. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\cite{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
$ P_\text{recognized}(t) = c \cdot \nu(t), $
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous mutations is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

### 2013-03-12 synmut.tex
remove packages disliked by plos

\pagestyle{myheadings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{eucal,eurosym,textcomp}
\usepackage[caption=false]{subfig}
\usepackage{pslatex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{flushleft}
{\Large
\textbf{\Title}
}
\\
Fabio~Zanini$^{1}$ and
Richard~A.~Neher$^{1,\ast}$
\\
\bf{1} \Affiliation
\\
$\ast$ E-mail: richard.neher@tuebingen.mpg.de
\end{flushleft}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Methods availability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All analysis and computer simulation scripts, as well as the sequence alignments
used, are available for download at \url{http://git.tuebingen.mpg.de/synmut}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Figure Legends}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{fig1.pdf}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e. the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{fig2.pdf}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows how quickly synonymous mutations are purged from the populations. 
Specifically, the figure shows the fraction of mutations that are still observed
after $\Delta t$ days, conditional on being observed in one of the three frequency 
intervals (different colors). 
In each frequency interval, the fraction of synonymous
mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
dashed horizontal lines.
Panel B) shows the fixation probability of derived synonymous
alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral mutations (indicated by the diagonal line).
This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{fig3.pdf}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
mutations fixed (red), sites at which mutations reached frequencies above 15\% but
were susequently lost (blue), and sites at which no mutations were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which mutations are lost.
Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\linewidth]{fig4.pdf}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix
\newpage
\appendix
\onecolumngrid
\setcounter{figure}{0}
\include{supplement_body}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\pagestyle{myheadings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{flushleft}
{\Large
\textbf{\Title}
}
\\
Fabio~Zanini$^{1}$ and
Richard~A.~Neher$^{1,\ast}$
\\
\bf{1} \Affiliation
\\
$\ast$ E-mail: richard.neher@tuebingen.mpg.de
\end{flushleft}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Methods availability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All analysis and computer simulation scripts, as well as the sequence alignments
used, are available for download at http://git.tuebingen.mpg.de/synmut.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Figure Legends}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=\linewidth]{fig1.pdf}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e. the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=\linewidth]{fig2.pdf}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows how quickly synonymous mutations are purged from the populations. 
Specifically, the figure shows the fraction of mutations that are still observed
after $\Delta t$ days, conditional on being observed in one of the three frequency 
intervals (different colors). 
In each frequency interval, the fraction of synonymous
mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
dashed horizontal lines.
Panel B) shows the fixation probability of derived synonymous
alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral mutations (indicated by the diagonal line).
This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=\linewidth]{fig3.pdf}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
mutations fixed (red), sites at which mutations reached frequencies above 15\% but
were susequently lost (blue), and sites at which no mutations were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which mutations are lost.
Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=\linewidth]{fig4.pdf}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix
%\newpage
%\appendix
%\setcounter{figure}{0}
%\include{supplement_body}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
